Exact Phylodynamics via Structured Markov Genealogy Processes

### 2023-07-31 ms.Rnw
good statements and proofs?

\paragraph{Likelihood conditional on history}

@@

Our second result concerns the likelihood of a given obscured genealogy conditional on the history.

@@

%% \AAK{Two approaches possible:}
%% \begin{enumerate}
%% \item Use the filter equations developed in the Appendix to compute $\Expect{\Indicator{\obs(\Prune_t)=\Vis_t}}$.
%%   In effect, we compute $\Prob{\Prune_t}$ and then sum over paintings $y$.
%% \item Change the order of the summation: $\Expect{\Indicator{\obs(\Prune_t)=\Vis_t}}=\Expect{\Expect{\Indicator{\obs(\Prune_t)=\Vis_t}\;\vert\;\Hist_t}}$.
%%   That is, conditional on $\Hist_t$, the likelihood of $\Vis_t$ is sum over all $\Prune_t$ such that $\obs(\Prune_t)=\Vis_t$.
%%   Therefore, any importance sampling scheme for $\Prune_t$ will do.
%%   In particular, we choose a scheme that paints the tree forward in time, driven by a mimic of $\X_t$.
%%   We show that this is equivalent to the specific filter equation.

%%   It may be useful in this to first show that $\Prob{\Prune_t|\Hist_t}$ can be expressed as an expectation over $y$, where $y$ is constrained to be equal $y^{\Prune_t}$.
%%   Then, show that the sum of the $\Prune_t$-constraint indicator functions is equal to the $\Vis_t$-constraint indicator function.
%% \end{enumerate}

%% \begin{figure}
%%   \begin{center}
%%     \input{figs/thm1diag}
%%   \end{center}
%%   \caption{\label{fig:thm1diag}}
%% \end{figure}

%% \begin{center}
%%   \input{figs/triangle1}
%% \end{center}

%% \begin{equation*}
%%   \frac{r!}{(r-s)!}\cdot\frac{(n-\ell)!}{n!}\cdot\frac{(n-r)!}{(n-r-\ell+s)!}
%%   =\frac{r!(n-r)!}{n!}\cdot\frac{(n-\ell)!}{(n-\ell-r+s)!(r-s)!}
%%   =\BinRatio{n}{\ell}{r}{s}
%% \end{equation*}

\paragraph{Integrating out the history: filter equations}

Next, we derive expressions for the likelihood of pruned and obscured genealogies, unconditional on the history.
For this, we use the filter equation technology developed in the Appendix.
In particular, the following theorem follows immediately from \cref{prop:fullfilt}.

The final result shows how to compute the likelihood of an obscured genealogy.

\begin{thm}
  Let $V^*_T$, $T>0$ be a given obscured genealogy.
  Let $\pi$, $\kappa_{V^*_T}$, and $\rdm{y}_t$ be as in \cref{thm:obsc_lik}.
  For $x,x'\in\Xspace$, $y,y'\in\Yspace$, $u\in\Jumps$, define
  \begin{equation*}
    \begin{gathered}
      \phi_u\big(t,x,y,y')=\BinRatio{n(x)}{\ell(y')}{r^u}{s(y,y')}\,\frac{\kappa_{V^*_T}(t,y,y')}{\pi_u(t,x,x',y,y')},\\
      \beta_u(t,x,x',y,y')=\alpha_u(t,x,x')\,\pi_u(t,x,x',y,y').
    \end{gathered}
  \end{equation*}
  Suppose $w=w(t,x)$ satisfies the filter equation
  \begin{equation}
    \begin{aligned}
      \frac{\partial{w}}{\partial{t}}(t,x,y)=
      &\sum_{uy'}{\int{w(t,x',y')\,\beta_u(t,x',x,y',y)\,\phi_u(t,x,y',y)\,\dd{x'}}}\\
      &\ -\sum_{uy'}{\int{w(t,x,y)\,\beta_u(t,x,x',y,y')\,\dd{x'}}}\\
      &\ +\sum_{e\in\event{P^*}{T}}{\delta(t,e)\,\left\{\sum_{uy'}{\int{w(t,x')\,\beta_u(t,x',x)\,\phi_u(t,x,y',y)\,\dd{x'}}}-w(t,x)\right\}},
    \end{aligned}
  \end{equation}
  with the initial condition $w(0,x,y)=p_0(x)\,\pi_0(y)$.
  Then the likelihood of $V^*_T$ is
  \begin{equation*}
    \lik(V^*_T)=\sum_y{\int{w(T,x,y)\,\dd{x}}}.
  \end{equation*}
\end{thm}

@@

\begin{defn}
  Suppose $\X_t$ is a continuous-time Markov process with Kolmogorov forward equation (KFE)
  \begin{equation}\label{eq:kfe2}
    \frac{\partial{w}}{\partial{t}}
    =\int{w(t,x')\,\beta(t,x',x)\,\dd{x'}}
    -\int{w(t,x)\,\beta(t,x,x')\,\dd{x'}}.
  \end{equation}
  Suppose that $B(t,x,x')>0$ and $\lambda(t,x)\in\R$ are given measurable functions.
  Let $\mathcal{M}\subseteq\Rp$ be countable such that $\mathcal{M}\cap[0,t]$ is finite for all $t$.
  Let $\nu>0$.
  We say that the equation
  \begin{equation}\label{eq:filtereq}
    \begin{aligned}
      \frac{\partial{w}}{\partial{t}}
      =&\int{w(t,x')\,\beta(t,x',x)\,B(t,x',x)\,\dd{x'}}
      -\int{w(t,x)\,\beta(t,x,x')\,\dd{x'}}
      -\lambda(t,x)\,w(t,x)\\
      &\quad+\sum_{e\in\mathcal{M}}\delta(t,e)\,\left\{\int{w(t,x')\,\frac{\beta(t,x',x)}{\nu}\,B(t,x',x)\,\dd{x'}}-w(t,x)\right\}
      +\nu\,w(t,x).
    \end{aligned}
  \end{equation}
  is the \emph{filter equation} with \emph{driver} $\X_t$, \emph{boost} $B$, \emph{decay} $\lambda$, and \emph{observed events} $\mathcal{M}$, \emph{relative to} $\nu$.
\end{defn}

\begin{remark}
  \Cref{eq:filtereq} is equivalent to
  \begin{equation*}
    \begin{gathered}
      \frac{\partial{w}}{\partial{t}}
      =\int{w(t,x')\,\beta(t,x',x)\,B(t,x',x)\,\dd{x'}}
      -\int{w(t,x)\,\beta(t,x,x')\,\dd{x'}}
      -(\lambda(t,x)-\nu)\,w(t,x),\quad t\notin{\mathcal{M}},\\
      w(t,x)=\int{\leftlim{w}(t,x')\,\frac{\beta(t,x',x)}{\nu}\,B(t,x',x)\,\dd{x'}},\quad t\in{\mathcal{M}}.
    \end{gathered}
  \end{equation*}
\end{remark}

@@

\begin{lemma}\label{lemma:filtereq}
  The filter equation \eqref{eq:filtereq} with $\mathcal{M}=\emptyset$ is satisfied by $w(t,x)=\int_0^{\infty}{v\,u(t,x,v)\,\dd{v}}$, where $u$ satisfies the KFE
  \begin{equation}\label{eq:filterlemma}
    \begin{aligned}
      \frac{\partial{u}}{\partial{t}}=&\int{u(t,x',v')\,\beta(t,x',x)\,\delta(v,B(t,x',x)\,v')\,\dd{x'}\,\dd{v'}}\\
      &\qquad-\int{u(t,x,v)\,\beta(t,x,x')\,\delta(v',B(t,x,x')\,v)\,\dd{x'}\,\dd{v'}}
      +\tfrac{\partial}{\partial{v}}\left[\lambda(t,x)\,v\,u(t,x,v)\right],
    \end{aligned}
  \end{equation}
  on the space $\Xspace\times(0,\infty)$.
  Here, $\delta(v,v')$ is the familiar Dirac $\delta$.
\end{lemma}
\begin{proof}
  \begin{equation*}
    \begin{aligned}
      \frac{\partial{w}}{\partial{t}}
      =&\int{v\,\frac{\partial{u}}{\partial{t}}(t,x,v)\,\dd{v}}\\
      =&\int{v\,u(t,x',v')\,\beta(t,x',x)\,\delta(v,B(t,x',x)v')\,\dd{v}\,\dd{x'}\,\dd{v'}}\\
      &\qquad-\int{v\,u(t,x,v)\,\beta(t,x,x')\,\delta(v',B(t,x,x')v)\,\dd{v}\,\dd{x'}\,\dd{v'}}\\
      &\qquad+\int{v\,\tfrac{\partial}{\partial{v}}\left[\lambda(t,x)\,v\,u(t,x,v)\right]\,\dd{v}}.\\
    \end{aligned}
  \end{equation*}
  Evaluating the first integral with respect to $v$, the second with respect to $v'$, and the third by parts, we obtain
  \begin{equation*}
    \begin{aligned}
      \frac{\partial{w}}{\partial{t}}
      =&\int{v'\,u(t,x',v')\,\beta(t,x',x)\,B(t,x',x)\,\dd{v'}\,\dd{x'}}
      -\int{v\,u(t,x,v)\,\beta(t,x,x')\,\dd{v}\,\dd{x'}}\\
      &\qquad-\lambda(t,x)\,\int{v\,u(t,x,v)\,\dd{v}},
    \end{aligned}
  \end{equation*}
  which is simplified to obtain \cref{eq:filtereq}.
\end{proof}

@@

\begin{prop}\label{prop:zeroboost}
  Suppose $\X_t$ is a non-explosive pure jump process with KFE
  \begin{equation*}
    \begin{gathered}
      \frac{\partial{w}}{\partial{t}}(t,x)=\int{w(t,x')\,\alpha(t,x',x)\,\dd{x'}}
      -\int{w(t,x)\,\alpha(t,x,x')\,\dd{x'}},\qquad
      w(0,x)=p_0(x).
    \end{gathered}
  \end{equation*}
  Let $\Hist_t$ be its history process.
  Suppose $\rdm{V}_t$ is a real-valued random process such that
  \begin{equation*}
    \Expect{\rdm{V}_t\;\vert\;\Hist_t=H_t}=\prod_{e\in\event{H}{t}}{Q(e,\leftlim{X}_e,X_e)\,B(e,\leftlim{X}_e,X_e)},
  \end{equation*}
  for some given measurable functions $B>{0}$, $Q\in\{0,1\}$.
  If $w$ satisfies the filter equation,
  \begin{equation*}
    \begin{aligned}
      \frac{\partial{w}}{\partial{t}}(t,x)
      =&\int{w(t,x')\,\alpha(t,x',x)\,Q(t,x',x)\,B(t,x',x)\,\dd{x'}}
      -\int{w(t,x)\,\alpha(t,x,x')\,Q(t,x,x')\,\dd{x'}}\\
      &\qquad-\int{w(t,x)\,\alpha(t,x,x')\,\Big(1-Q(t,x,x')\Big)\,\dd{x'}},
    \end{aligned}
  \end{equation*}
  then $\Expect{\rdm{V}_t}=\int{w(t,x)\,\dd{x}}$.
\end{prop}
\begin{proof}
  Apply \cref{lemma:filtereq} with $\beta(t,x,x')=\alpha(t,x,x')\,Q(t,x,x')$.
\end{proof}

@@

Filter equations allow us to pass easily between equivalent representations of a process.
For example, an equivalent way of representing $\X_t$ is in terms of its embedded chain and event times.
Let $\rdm{\hat{X}}_k$ be the embedded chain of $\X_t$ and let $\rdm{\hat{T}}_k$ be the point process of its event times.
It is elementary that
\fontsize{10pt}{12pt}\selectfont
\begin{equation*}
  \begin{gathered}
    \Prob{\rdm{\hat{X}}_k=\hat{X}_k\;\Big\vert\;\rdm{\hat{X}}_{k-1}=\hat{X}_{k-1},\rdm{\hat{T}}_k=\hat{T}_k}=\frac{\alpha(\hat{T}_k,\hat{X}_{k-1},\hat{X}_k)}{\int{\alpha(\hat{T}_k,\hat{X}_{k-1},x')\dd{x'}}},\\
    \Prob{\rdm{\hat{T}}_k>\rdm{\hat{T}}_{k-1}+t\;\Big\vert\;\rdm{\hat{X}}_{k-1}=\hat{X}_{k-1},\rdm{\hat{T}}_{k-1}=\hat{T}_{k-1}}=\exp{\left(-\int_{0}^{t}{\int{\alpha(\hat{T}_{k-1}+s,\hat{X}_{k-1},x')\,\dd{x'}}\,\dd{s}}\right)}.
  \end{gathered}
\end{equation*}
\normalfont
Making the definitions,
\begin{equation}\label[pluralequation]{eq:nudefs}
  \begin{gathered}
    A(t,x)=\int{\alpha(t,x,x')\,\dd{x'}}, \qquad
    \pi(t,x,x')=\frac{\alpha(t,x,x')}{A(t,x)},\\
    B(t,x,x')=\frac{A(t,x)}{\nu}, \qquad
    \lambda(t,x)=A(t,x)-\nu,
  \end{gathered}
\end{equation}
we can rewrite the KFE as
\begin{equation}\label{eq:poissdriver}
  \frac{\partial{w}}{\partial{t}}(t,x)=
  \int{w(t,x')\,\nu\,\pi(t,x',x)\,B(t,x,x')\,\dd{x'}}
  -\int{w(t,x)\,\nu\,\pi(t,x,x')\,\dd{x'}}
  -\lambda(t,x)\,w(t,x).
\end{equation}
Here, $\nu>0$ is the intensity of a time-homogenous Poisson process.
Note that $\pi$ is the probability kernel of the embedded chain $\rdm{\hat{X}}$ and $A(t,x)$ is the intensity of the $\rdm{\hat{T}}_k$ process.
We recognize this equation as the filter equation with boost $B$, decay $\lambda$, and driver generated by $\nu\,\pi(t,x,x')$.
It corresponds to the following procedure for simulating $\X_t$:
\begin{compactenum}[(a)]
\item Simulate jump times according to the rate-$\nu$ Poisson process.
\item Simulate the embedded chain $\rdm{\hat{X}}_k$ using the kernel $\pi$.
\item Weight the realization by the product of the $B$ factors.
  Note that this makes the appropriate importance-sampling correction.
\end{compactenum}

By integrating equations of this kind, we effectively compute expectations with respect to the $\X_t$ process.
In particular, we can use \cref{prop:zeroboost} or \cref{eq:poissdriver} to integrate out the unobserved event times $\rdm{\hat{T}}_k$, as well as the unobserved states $\rdm{\hat{X}}_k$.
However, we also need to be able to compute likelihoods for data that include observed event times.
Technically, in such a case, the likelihood is a density on the cone $\mathbb{C}_t=\coprod_{m=0}^{\infty}{\mathbb{C}^m_t}$, where $\mathbb{C}^m_t=\{t\in\halfclosed{0,t}\;\vert\;t_1<t_2<\dots<t_m\}$.
To avoid having to work directly with such complex objects, it is convenient to represent the likelihood as a density with respect to a constant-rate Poisson point process.
This is accomplished as follows.

Suppose the finite set of observed event times $\mathcal{M}\subseteq\halfclosed{0,t}$ is fixed and $|\mathcal{M}|=M$.
Let $\nu>0$ be a constant rate and consider the mixed point process with intensity $\nu+\sum_{e\in\mathcal{M}}\delta(t,e)$.
This point process has events at $e\in\mathcal{M}$ with probability 1 as well as additional, Poisson-distributed random event times.
Using this as the driver in \cref{eq:poissdriver}, we obtain the filter equation
\begin{equation*}
  \begin{aligned}
    \frac{\partial{w}}{\partial{t}}(t,x)=&\int{w(t,x')\,\nu\,\pi(t,x',x)\,B(t,x',x)\,\dd{x'}}-
    \int{w(t,x)\,\nu\,\pi(t,x,x')\,\dd{x'}}
    -\lambda(t,x)\,w(t,x)\\
    &\quad+\sum_{e\in\mathcal{M}}\delta(t,e)\,\left\{\int{w(t,x')\,\pi(t,x',x)\,B(t,x,x')\,\dd{x'}}
    -\int{w(t,x)\,\pi(t,x,x')\,\dd{x'}}\right\}.\\
  \end{aligned}
\end{equation*}
This is equivalent to
\begin{equation}\label[pluralequation]{eq:mixedKFE}
  \begin{aligned}
    \frac{\partial{w}}{\partial{t}}(t,x)
    =&\int{w(t,x')\,\alpha(t,x',x)\,\dd{x'}}
    -\int{w(t,x)\,\alpha(t,x,x')\,\dd{x'}},\quad t\notin{\mathcal{M}},\\
    w(e,x)=&\int{\leftlim{w}(e,x')\,\pi(e,x',x)\,B(e,x',x)\,\dd{x'}}
    =\tfrac{1}{\nu}\,\int{\leftlim{w}(e,x')\,\alpha(e,x',x)\,\dd{x'}},\quad e\in\mathcal{M}.
  \end{aligned}
\end{equation}
\Cref{eq:mixedKFE} is a filter equation relative to $\nu$ with observed events $\mathcal{M}$.
It defines a probability measure $\mu_M$ on $\Xspace\times{\mathbb{C}^M_t}$.
Specifically, $w(t,x)$ is the Radon-Nikodym derivative of $\mu_M$ with respect to the product of the measure on $\Xspace$ and the probability measure of the rate-$\nu$ Poisson process on $\mathbb{C}^M_t$.
That is,
\begin{equation*}
  \mu^M_t(\dd{x}\,\dd{t_1}\,\cdots\,\dd{t_M})=\nu^{M}\,e^{-\nu\,t}\,w(t,x)\,\dd{x}\,\dd{t_1}\,\cdots\,\dd{t_M}.
\end{equation*}

%% We can re-express the history-process density (\cref{eq:Hdens}) using a filter equation as well.
%% Specifically, one can readily verify that if $H_t$ is a history and $w=w(t,x,u)$ satisfies
%% \begin{equation}\label{eq:Hdens_filt}
%%   \begin{aligned}
%%     \frac{\partial{w}}{\partial{t}}(t,x,u)=&-w(t,x,u)\,\sum_{u'}{\int{\alpha_{u'}(t,x,x')\,\dd{x'}}}\\
%%     &\quad+\sum_{e\in\event{H}{t}}{\delta(t,e)\,\left\{\sum_{u'}{\int{w(t,x',u')\,\frac{\alpha_{u'}(t,x',x)}{\nu}\,\dd{x'}}}-w(t,x,u)\right\}},
%%   \end{aligned}
%% \end{equation}
%% then $\pi^H(\dd{t_1}\cdots\dd{t_K})=\nu^K\,e^{-\nu\,t}\,\sum_u{\int{w(t,x,u)\,\dd{x'}}}\,\dd{t_1}\cdots\dd{t_K}$,
%% where $K=\left\vert\event{H}{t}\right\vert$.
%% \Cref{eq:Hdens_filt} is the filter equation, relative to $\nu$, with null driver, boost $B(t,x,x',u,u')=\alpha_{u'}(t,x,x')$, and decay $\lambda(t,x,u)=\sum_{u'}{\int{\alpha_{u'}(t,x,x')\,\dd{x'}}}$.

The same kind of reasoning shows how to compute expectations of a real-valued process $\rdm{V}_t$ directed by $\X_t$.
As before, we will express the expectation $\Expect{\rdm{V}_t}$ as a density with respect to the rate-$\nu$ Poisson point process.

>>>>

\paragraph{Likelihood for pruned genealogies}

@@

Next, we show how the likelihood of a pruned genealogies, unconditional on the history, can be computed.
For this, we use the filter equation technology developed in the Appendix.
In particular, the following theorem follows immediately from \cref{prop:fullfilt}.

\paragraph{Likelihood for obscured genealogies}

Our next result concerns the likelihood of a given obscured genealogy conditional on the history.

@@

The final result shows how to compute the likelihood of an obscured genealogy.
It is an immediate consequence of \cref{thm:obsc_lik,prop:fullfilt}.

\begin{thm}
  Let $V^*_T$, $T>0$ be a given obscured genealogy.
  Let $\pi$, $\kappa_{V^*_T}$, and $\rdm{y}_t$ be as in \cref{thm:obsc_lik}.
  For $x,x'\in\Xspace$, $y,y'\in\Yspace$, $u\in\Jumps$, define
  \begin{equation*}
    \begin{gathered}
      \phi_u\big(t,x,y,y')=\BinRatio{n(x)}{\ell(y')}{r^u}{s(y,y')}\,\frac{\kappa_{V^*_T}(t,y,y')}{\pi_u(t,x,x',y,y')},\\
      \beta_u(t,x,x',y,y')=\alpha_u(t,x,x')\,\pi_u(t,x,x',y,y').
    \end{gathered}
  \end{equation*}
  Suppose $w=w(t,x)$ satisfies the filter equation
  \begin{equation}
    \begin{aligned}
      \frac{\partial{w}}{\partial{t}}(t,x,y)=
      &\sum_{uy'}{\int{w(t,x',y')\,\beta_u(t,x',x,y',y)\,\phi_u(t,x,y',y)\,\dd{x'}}}\\
      &\ -\sum_{uy'}{\int{w(t,x,y)\,\beta_u(t,x,x',y,y')\,\dd{x'}}}\\
      &\ +\sum_{e\in\event{V^*}{T}}{\delta(t,e)\,\left\{\sum_{uy'}{\int{w(t,x')\,\beta_u(t,x',x)\,\phi_u(t,x,y',y)\,\dd{x'}}}-w(t,x)\right\}},
    \end{aligned}
  \end{equation}
  with the initial condition $w(0,x,y)=p_0(x)\,\pi_0(y)$.
  Then the likelihood of $V^*_T$ is
  \begin{equation*}
    \lik(V^*_T)=\sum_y{\int{w(T,x,y)\,\dd{x}}}.
  \end{equation*}
\end{thm}

@@

\begin{defn}
  Suppose $\X_t$ is a continuous-time Markov process with Kolmogorov forward equation (KFE)
  \begin{equation}\label{eq:kfe2}
    \frac{\partial{w}}{\partial{t}}
    =\int{w(t,x')\,\beta(t,x',x)\,\dd{x'}}
    -\int{w(t,x)\,\beta(t,x,x')\,\dd{x'}}.
  \end{equation}
  Suppose that $B(t,x,x')\ge{0}$ and $\lambda(t,x)\in\R$ are given measurable functions.
  Let $\mathcal{M}\subseteq\Rp$ be countable such that $\mathcal{M}\cap[0,t]$ is finite for all $t$.
  We say that the equation
  \begin{equation}\label{eq:filtereq}
    \begin{aligned}
      \frac{\partial{w}}{\partial{t}}
      =&\int{w(t,x')\,\beta(t,x',x)\,B(t,x',x)\,\dd{x'}}
      -\int{w(t,x)\,\beta(t,x,x')\,\dd{x'}}
      -\lambda(t,x)\,w(t,x)\\
      &\quad+\sum_{e\in\mathcal{M}}\delta(t,e)\,\left\{\int{w(t,x')\,\beta(t,x',x)\,B(t,x',x)\,\dd{x'}}-w(t,x)\right\}.
    \end{aligned}
  \end{equation}
  is the \emph{filter equation} with \emph{driver} $\X_t$ (or \emph{generated by} $\beta$), \emph{boost} $B$, \emph{decay} $\lambda$, and \emph{observed events} $\mathcal{M}$.
\end{defn}

\begin{remark}\label{rem:equivfilt}
  \Cref{eq:filtereq} is equivalent to
  \begin{equation*}
    \begin{gathered}
      \frac{\partial{w}}{\partial{t}}
      =\int{w(t,x')\,\beta(t,x',x)\,B(t,x',x)\,\dd{x'}}
      -\int{w(t,x)\,\beta(t,x,x')\,\dd{x'}}
      -\lambda(t,x)\,w(t,x),\quad t\notin{\mathcal{M}},\\
      w(t,x)=\int{\leftlim{w}(t,x')\,\beta(t,x',x)\,B(t,x',x)\,\dd{x'}},\quad t\in{\mathcal{M}}.
    \end{gathered}
  \end{equation*}
\end{remark}

@@

\begin{lemma}\label{lemma:filtereq}
  The filter equation \eqref{eq:filtereq} with $\mathcal{M}=\emptyset$ and $B>0$ is satisfied by $w(t,x)=\int_0^{\infty}{v\,u(t,x,v)\,\dd{v}}$, where $u$ satisfies the KFE
  \begin{equation}\label{eq:filterlemma}
    \begin{aligned}
      \frac{\partial{u}}{\partial{t}}=&\int{u(t,x',v')\,\beta(t,x',x)\,\delta(v,B(t,x',x)\,v')\,\dd{x'}\,\dd{v'}}\\
      &\qquad-\int{u(t,x,v)\,\beta(t,x,x')\,\delta(v',B(t,x,x')\,v)\,\dd{x'}\,\dd{v'}}
      +\tfrac{\partial}{\partial{v}}\left[\lambda(t,x)\,v\,u(t,x,v)\right],
    \end{aligned}
  \end{equation}
  on the space $\Xspace\times(0,\infty)$.
  Here, $\delta(v,v')$ is the familiar Dirac $\delta$.
\end{lemma}
\begin{proof}
  \begin{equation*}
    \begin{aligned}
      \frac{\partial{w}}{\partial{t}}
      =&\int{v\,\frac{\partial{u}}{\partial{t}}(t,x,v)\,\dd{v}}\\
      =&\int{v\,u(t,x',v')\,\beta(t,x',x)\,\delta(v,B(t,x',x)v')\,\dd{v}\,\dd{x'}\,\dd{v'}}\\
      &\qquad-\int{v\,u(t,x,v)\,\beta(t,x,x')\,\delta(v',B(t,x,x')v)\,\dd{v}\,\dd{x'}\,\dd{v'}}\\
      &\qquad+\int{v\,\tfrac{\partial}{\partial{v}}\left[\lambda(t,x)\,v\,u(t,x,v)\right]\,\dd{v}}.\\
    \end{aligned}
  \end{equation*}
  Evaluating the first integral with respect to $v$, the second with respect to $v'$, and the third by parts, we obtain
  \begin{equation*}
    \begin{aligned}
      \frac{\partial{w}}{\partial{t}}
      =&\int{v'\,u(t,x',v')\,\beta(t,x',x)\,B(t,x',x)\,\dd{v'}\,\dd{x'}}
      -\int{v\,u(t,x,v)\,\beta(t,x,x')\,\dd{v}\,\dd{x'}}\\
      &\qquad-\lambda(t,x)\,\int{v\,u(t,x,v)\,\dd{v}},
    \end{aligned}
  \end{equation*}
  which is simplified to obtain \cref{eq:filtereq}.
\end{proof}

@@

\begin{prop}\label{prop:zeroboost}
  Suppose $\X_t$ is a non-explosive pure jump process with KFE
  \begin{equation*}
    \begin{gathered}
      \frac{\partial{w}}{\partial{t}}(t,x)=\int{w(t,x')\,\alpha(t,x',x)\,\dd{x'}}
      -\int{w(t,x)\,\alpha(t,x,x')\,\dd{x'}},\qquad
      w(0,x)=p_0(x).
    \end{gathered}
  \end{equation*}
  Let $\Hist_t$ be its history process.
  Suppose $\rdm{V}_t$ is a real-valued random process such that
  \begin{equation}\label{eq:Vprod}
    \Expect{\rdm{V}_t\;\vert\;\Hist_t=H_t}=\prod_{e\in\event{H}{t}}{B(e,\leftlim{X}_e,X_e)},
  \end{equation}
  for some given measurable function $B\ge{0}$.
  If $w$ satisfies the filter equation,
  \begin{equation*}
    \frac{\partial{w}}{\partial{t}}(t,x)
    =\int{w(t,x')\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}}
    -\int{w(t,x)\,\alpha(t,x,x')\,\dd{x'}}\\
  \end{equation*}
  then $\Expect{\rdm{V}_t}=\int{w(t,x)\,\dd{x}}$.
\end{prop}
\begin{proof}
  Define $Q(t,x,x')=\Indicator{B(t,x,x')>0}$ and apply \cref{lemma:filtereq} with
  \begin{equation*}
    \begin{gathered}
      \beta(t,x,x')=\alpha(t,x,x')\,Q(t,x,x'),
      \qquad
      \lambda=\int{\alpha(t,x,x')\,\left(1-Q(t,x,x')\right)\dd{x'}}.
    \end{gathered}
  \end{equation*}
\end{proof}

@@

The following result extends \cref{prop:zeroboost} to the situation where $\mathcal{M}\ne\emptyset$, \ie where the times of some of the events in the history are observed.

\begin{prop}\label{prop:fullfilt}
  Let $\X_t$, $\Hist_t$, $B$, and $\rdm{V}_t$ be as in \cref{prop:zeroboost}.
  Suppose that $\mathcal{M}\subseteq\halfclosed{0,t}$ is a fixed, finite set of observed event times.
  Let $\Expect{\rdm{V}_t\;\vert\;\mathcal{M}}$ denote the density of $\rdm{V}_t$ conditional on $\mathcal{M}\subseteq\event{\Hist}{t}$.
  If $w(t,x)$ satisfies $w(0,x)=p_0(x)$ and the filter equation
  \begin{equation*}
    \begin{aligned}
      \frac{\partial{w}}{\partial{t}}
      =&\int{w(t,x')\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}}
      -\int{w(t,x)\,\alpha(t,x,x')\,\dd{x'}}\\
      &\quad+\sum_{e\in\mathcal{M}}\delta(t,e)\,\left\{\int{w(t,x')\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}}-w(t,x)\right\},
    \end{aligned}
  \end{equation*}
  then $\Expect{\rdm{V}_t\;\vert\;\mathcal{M}}=\int{w(t,x)\,\dd{x}}$.
\end{prop}
\begin{proof}
  Suppose $s\in[0,t]$ and let
  \begin{equation*}
    w(s,x)=\Expect{\prod_{e\in\event{\Hist}{s}}{B\left(e,\Xt_e,\X_e\right)}}.
  \end{equation*}
  That is, $w(s,x)$ is the expectation (over histories) of the partial products of \cref{eq:Vprod}.
  If $s\notin{\mathcal{M}}$, then for $\Delta>0$ sufficiently small,
  \begin{equation*}
    \begin{aligned}
      w(s,x)=&\left(1-\Delta\,\int{\alpha(s-\Delta,x,x')\,\dd{x'}}\right)\,w(s-\Delta,x)\\
      &\qquad+\int{\alpha(s-\Delta,x',x)\,B(s-\Delta,x',x)\,w(s-\Delta,x')\,\dd{x'}}+o(\Delta).
    \end{aligned}
  \end{equation*}
  Taking the limit as $\Delta\downarrow{0}$, we obtain
  \begin{equation*}
    \frac{\partial{w}}{\partial{s}}(s,x)
    =\int{w(s,x')\,\alpha(s,x',x)\,B(s,x',x)\,\dd{x'}}
    -\int{w(s,x)\,\alpha(s,x,x')\,\dd{x'}}.
  \end{equation*}
  If $s\in\mathcal{M}$, then
  \begin{equation*}
    w(s,x)=\int{\wt(s,x')\,\alpha(s,x',x)\,B(s,x',x)\,\dd{x'}}.
  \end{equation*}
  By \cref{rem:equivfilt}, the result follows.
\end{proof}

%% Filter equations allow us to pass easily between equivalent representations of a process.
%% For example, an equivalent way of representing $\X_t$ is in terms of its embedded chain and event times.
%% Let $\rdm{\hat{X}}_k$ be the embedded chain of $\X_t$ and let $\rdm{\hat{T}}_k$ be the point process of its event times.
%% It is elementary that
%% \fontsize{10pt}{12pt}\selectfont
%% \begin{equation*}
%%   \begin{gathered}
%%     \Prob{\rdm{\hat{X}}_k=\hat{X}_k\;\Big\vert\;\rdm{\hat{X}}_{k-1}=\hat{X}_{k-1},\rdm{\hat{T}}_k=\hat{T}_k}=\frac{\alpha(\hat{T}_k,\hat{X}_{k-1},\hat{X}_k)}{\int{\alpha(\hat{T}_k,\hat{X}_{k-1},x')\dd{x'}}},\\
%%     \Prob{\rdm{\hat{T}}_k>\rdm{\hat{T}}_{k-1}+t\;\Big\vert\;\rdm{\hat{X}}_{k-1}=\hat{X}_{k-1},\rdm{\hat{T}}_{k-1}=\hat{T}_{k-1}}=\exp{\left(-\int_{0}^{t}{\int{\alpha(\hat{T}_{k-1}+s,\hat{X}_{k-1},x')\,\dd{x'}}\,\dd{s}}\right)}.
%%   \end{gathered}
%% \end{equation*}
%% \normalfont
%% Fixing $\nu>0$ and making the definitions,
%% \begin{equation}\label[pluralequation]{eq:nudefs}
%%   \begin{gathered}
%%     A(t,x)=\int{\alpha(t,x,x')\,\dd{x'}}, \qquad
%%     \pi(t,x,x')=\frac{\alpha(t,x,x')}{A(t,x)},\\
%%     B(t,x,x')=\frac{A(t,x)}{\nu}, \qquad
%%     \lambda(t,x)=A(t,x)-\nu,
%%   \end{gathered}
%% \end{equation}
%% we can rewrite the KFE as
%% \begin{equation}\label{eq:poissdriver}
%%   \frac{\partial{w}}{\partial{t}}(t,x)=
%%   \int{w(t,x')\,\nu\,\pi(t,x',x)\,B(t,x,x')\,\dd{x'}}
%%   -\int{w(t,x)\,\nu\,\pi(t,x,x')\,\dd{x'}}
%%   -\lambda(t,x)\,w(t,x).
%% \end{equation}
%% Here, $\nu$ is the intensity of a time-homogenous Poisson process.
%% Note that $\pi$ is the probability kernel of the embedded chain $\rdm{\hat{X}}$ and $A(t,x)$ is the intensity of the $\rdm{\hat{T}}_k$ process.
%% We recognize this equation as the filter equation with boost $B$, decay $\lambda$, and driver generated by $\nu\,\pi(t,x,x')$.
%% It corresponds to the following procedure for simulating $\X_t$:
%% \begin{compactenum}[(a)]
%% \item Simulate jump times according to the rate-$\nu$ Poisson process.
%% \item Simulate the embedded chain $\rdm{\hat{X}}_k$ using the kernel $\pi$.
%% \item Weight the realization by the product of the $B$ factors.
%%   Note that this makes the appropriate importance-sampling correction.
%% \end{compactenum}
